A Unified Treatment of Tractability for \\ Approximation Problems Defined on Hilbert Spaces

### 2023-11-15 Older_versions/GeneralizedTractability2023August10.tex
Updates from Overleaf

The remainder of this paper is organized as follows. In Section \ref{sec:spt}, we give the definitions of the considered tractability notions, in particular (strong) tractability. We consider strong tractability when there is an upper bound on the information complexity that is independent of the dimension, $d$. We then turn to tractability  when the bound on the information complexity may depend on both $d$ and $\varepsilon^{-1}$. In Section \ref{sec:examples} we provide examples of tractability functions for both the algebraic and exponential cases. We also consider other notions of tractability such as quasi-polynomial tractability, and introduce other notions of tractability. A generalized notion of weak tractability is given in Section \ref{sec:subh}. In Section \ref{sec:restricted}, we consider \peter{tractability on} a restricted domain .

@@

For this definition of tractability to make sense we assume that
\begin{subequations} \label{eq:Tconditions}
\begin{equation} \label{eq:Tnondecreasing}
	T \text{ is non-decreasing in all variables,}
\end{equation}
which implies that the problem is expected to be no easier by decreasing the tolerance, $\varepsilon$, or increasing the dimension, $d$\footnote{This is why we make $T$ a function of $\varepsilon^{-1}$ rather than of $\varepsilon$.} Furthermore, increasing $\vp$ allows for a possibly looser bound on the information complexity.  Since there are an infinite number of positive \peter{singular values}, it also makes sense to assume that
\begin{equation}\label{eq:Tlim}
	\lim_{\varepsilon \to 0} T(\varepsilon^{-1},d,\vp) = \infty \qquad \forall d \in \naturals, \ \vp \in [0,\infty)^s.
\end{equation}
Since $T(\cdot,d,\vp)$ is non-decreasing, we may define the following limit, which we assume will always be positive:
\begin{equation}\label{eq:T0}	T(0,d,\vp):=\lim_{\varepsilon\to\infty}T(\varepsilon^{-1},d,\vp) = \inf_\varepsilon T(\varepsilon^{-1},d,\vp) \ge T(0,1,\vzero) > 0.
\end{equation}
\end{subequations}
Finally, there is a technical assumption required for our analysis in \Cref{sec:spt,sec:restricted}.  There exists a $K_{\vp,\tau}$ depending on $\vp$ and $\tau$, but  independent of $\varepsilon$ and $d$, such that
\begin{equation} \label{eq:ptauassume}
	[T(\varepsilon^{-1},d,\vp)]^\tau \le K_{\vp,\tau} T(\varepsilon^{-1},d,\tau \vp)   \quad \forall \varepsilon \in (0,\infty), \ d \in \naturals, \ \vp\in[0,\infty)^s, \ \tau\in [1,\infty).
\end{equation}

@@

\peter{The following proposition states that the conditions \eqref{eq:exp_spt} and \eqref{eq:KW_exp_spt}, as well as the conditions \eqref{eq:exp_pt} and \eqref{eq:KW_exp_pt} are indeed equivalent. 
The proof of this result is deferred to the appendix, where we will prove this statement for the latter two conditions, i.e., for the conditions on exponential polynomial tractability. The case of exponential strong polynomial tractability is not explicitly included, as it is actually simpler and can be treated analogously.
\begin{proposition}\label{prop:equiv_conditions}
The conditions \eqref{eq:exp_spt} and \eqref{eq:KW_exp_spt} are equivalent. Furthermore, the conditions \eqref{eq:exp_pt} and \eqref{eq:KW_exp_pt} are equivalent.
\end{proposition}
}
In the following example, we provide a brief summary of the above examples of tractability. We would like to emphasize that this example was previously known (see, e.g., \cite{GW06}) where in this case the roles of $\varepsilon^{-1}$ and $d$ are separated. In order to accomplish this, we will first define two functions $\phi$ and $\psi$ with positive values that are increasing in both of their arguments which will allow to take into account a variety of possible exponents of $\varepsilon^{-1}$ and $d$ \peter{[Is this really what we would like to express in this context?]}. The equivalent conditions for this example take the same form as those in Theorems \ref{thm_main_strong_tract2} and \ref{thm_main_tract2}.

@@

\section{Discussion and Summary}

We have shown that a number of proofs of equivalent conditions for (strong) tractability of various types can be unified into a handful of proofs.  Even the arguments underlying these handful of proofs are quite similar and could be consolidated even further.  However some questions remain for further work:
\begin{itemize}
    \item What interesting cases of $T$ do not correspond to known tractability measures? 
    \item What interesting cases are there where there is no (strong) tractability on an unrestricted domain, but there is (strong) tractability on a nontrivial restricted domain, such as $(0,\infty) \times \{1, \ldots, d_{\max}] \cup (\varepsilon_{\min}^{-1},\infty) \times \naturals$?
    \item For Hilbert spaces and solution operators, $\{\SOL_d : \cf_d \to \cg_d\}_{d \in \naturals}$, of tensor product form, how can the equivalent conditions for (strong) tractability be simplified, even when $T(a_1 \cdots a_d,d,\vp) \ne T_1(a_1,d,\vp)\cdots T_1(a_d,d,\vp)$ for some $T_1$?
    \item For information complexity defined over cones of functions rather than balls (see e.g., \cite{DinHic20a,DinEtal20a}), what can be said about equivalent conditions for (strong) tractability?
\end{itemize}

>>>>

The remainder of this paper is organized as follows. In Section \ref{sec:spt}, we give the definitions of the considered tractability notions, in particular (strong) tractability. We consider strong tractability when there is an upper bound on the information complexity that is independent of the dimension, $d$. We then turn to tractability  when the bound on the information complexity may depend on both $d$ and $\varepsilon^{-1}$. In Section \ref{sec:examples} we provide examples of tractability functions for both the algebraic and exponential cases. We also consider other notions of tractability such as quasi-polynomial tractability, \ and introduce other notions of tractability. A generalized notion of weak tractability is given in Section \ref{sec:subh}. In Section \ref{sec:restricted}, we consider \peter{tractability on} a restricted domain.

@@

For this definition of tractability to make sense we assume that
\begin{subequations} \label{eq:Tconditions}
\begin{equation} \label{eq:Tnondecreasing}
	T \text{ is non-decreasing in all variables,}
\end{equation}
which implies that the problem is expected to be no easier by decreasing the tolerance, $\varepsilon$, or increasing the dimension, $d$\footnote{This is why we make $T$ a function of $\varepsilon^{-1}$ rather than of $\varepsilon$.}. Furthermore, increasing $\vp$ allows for a possibly looser bound on the information complexity. Since there are an infinite number of positive \peter{singular values}, it also makes sense to assume that
\begin{equation}\label{eq:Tlim}
	\lim_{\varepsilon \to 0} T(\varepsilon^{-1},d,\vp) = \infty \qquad \forall d \in \naturals, \ \vp \in [0,\infty)^s.
\end{equation}
Since $T(\cdot,d,\vp)$ is non-decreasing, we may define the following limit, which we assume will always be positive:
\begin{equation}\label{eq:T0}	T(0,d,\vp):=\lim_{\varepsilon\to\infty}T(\varepsilon^{-1},d,\vp) = \inf_\varepsilon T(\varepsilon^{-1},d,\vp) \ge T(0,1,\vzero) > 0.
\end{equation}
\end{subequations}
Finally, there is a technical assumption required for our analysis in \Cref{sec:spt,sec:restricted}.  There exists a $K_{\vp,\tau}$ depending on $\vp$ and $\tau$, but  independent of $\varepsilon$ and $d$, such that
\begin{equation} \label{eq:ptauassume}
	[T(\varepsilon^{-1},d,\vp)]^\tau \le K_{\vp,\tau} T(\varepsilon^{-1},d,\tau \vp)   \quad \forall \varepsilon \in (0,\infty), \ d \in \naturals, \ \vp\in[0,\infty)^s, \ \tau\in [1,\infty).
\end{equation}

@@

\peter{The following proposition states that the conditions \eqref{eq:exp_spt} and \eqref{eq:KW_exp_spt}, as well as the conditions \eqref{eq:exp_pt} and \eqref{eq:KW_exp_pt} are indeed equivalent. 
The proof of this result is deferred to the appendix, where we will prove this statement for the latter two conditions, i.e., for the conditions on exponential polynomial tractability. The case of exponential strong polynomial tractability is not explicitly included, as it is actually simpler and can be treated analogously.
\begin{proposition}\label{prop:equiv_conditions}
The conditions \eqref{eq:exp_spt} and \eqref{eq:KW_exp_spt} are equivalent. Furthermore, the conditions \eqref{eq:exp_pt} and \eqref{eq:KW_exp_pt} are equivalent.
\end{proposition}
}
In the following example, we provide a brief summary of the above examples of tractability. We would like to emphasize that this example was previously known (see, e.g., \cite{GW06}) where in this case the roles of $\varepsilon^{-1}$ and $d$ are separated. In order to accomplish this, we will first define two functions $\phi$ and $\psi$ with positive values that are increasing in both of their arguments. The equivalent conditions for this example take the same form as those in Theorems \ref{thm_main_strong_tract2} and \ref{thm_main_tract2}.

@@

\section{Discussion and Further Work}

We have shown that many proofs of equivalent conditions for (strong) tractability of various types can be unified into a handful of proofs.  Even the arguments underlying these handful are quite similar and could be consolidated even further.  This unification simplifies our understanding of these equivalent conditions and spotlights the key ideas about what makes a class of problems hard.

However, some questions remain:
\begin{itemize}
    \item What interesting cases of $T$ do not correspond to known tractability measures? 
    \item What interesting cases are there where there is no (strong) tractability on an unrestricted domain, but there is (strong) tractability on a nontrivial restricted domain, such as $(0,\infty) \times \{1, \ldots, d_{\max}] \cup (\varepsilon_{\min}^{-1},\infty) \times \naturals$?
     \item For Hilbert spaces and solution operators, $\{\SOL_d : \cf_d \to \cg_d\}_{d \in \naturals}$, of tensor product form, the singular values take the form $\lambda_{i,d} = \lambda_{i,d,1} \cdots  \lambda_{i,d,d}$.  What more can be said about the  equivalent conditions  for (strong) tractability in this case?
    \item For information complexity defined over cones of functions rather than balls (see e.g., \cite{DinHic20a,DinEtal20a}), what can be said about equivalent conditions for (strong) tractability?
\end{itemize}

### 2023-11-15 arxiv_copy.tex
Updates from Overleaf

\documentclass[sort&compress]{elsarticle}
\usepackage{latexsym,amsfonts,amsmath,graphics,mathtools,amssymb,booktabs,xspace}
\usepackage{epsfig}
\usepackage[notref,notcite]{showkeys}
%\usepackage[active]{srcltx}
%\usepackage{fdsymbol} %removing this package solves the problem with \mathcal{I} not showing up properly.
\usepackage{enumitem}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[normalem]{ulem}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{mathrsfs}
\usepackage{cleveref}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newdefinition{definition}{Definition}
\newdefinition{example}{Example}
\newproof{proof}{Proof}

@@

\author[1]{Onyekachi Emenike}
\ead{kachi.emenike12@gmail.com}
\author[1]{Fred J. Hickernell}
\ead{hickernell@iit.edu}
\author[2]{Peter Kritzer}
\ead{peter.kritzer@oeaw.ac.at}

\date{\today}
\affiliation[1]{organization={Department of Applied Mathematics, Illinois Institute of Technology},
addressline={10 West 32${}^{\text{nd}}$ Street, Room 220},
postcode={60616},
city={Chicago, IL},
country={United States}}
\affiliation[2]{organization={Johann Radon Institute for Computational and Applied Mathematics (RICAM), Austrian Academy of Sciences},
addressline={Altenbergerstr. 69},
postcode={4040},
city={Linz},
country={Austria}}

@@

\maketitle

@@

We give the proof of Proposition \ref{prop:equiv_conditions} here.
\begin{proof}[Proposition 1]
We will give the proof of equivalence of the conditions \eqref{eq:exp_pt} and \eqref{eq:KW_exp_pt}. The proof of the equivalence of \eqref{eq:exp_spt} and \eqref{eq:KW_exp_spt} is analogous and simpler (actually, the latter proof can be seen as a special case of the former, by assuming all exponents of $d$ to be zero).

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage{latexsym,amsfonts,amsmath,graphics,mathtools,amssymb,amsthm,booktabs}
\usepackage{epsfig}
\usepackage[notref,notcite]{showkeys}
%\usepackage[active]{srcltx}
%\usepackage{fdsymbol} %removing this package solves the problem with \mathcal{I} not showing up properly.
\usepackage{enumitem}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[normalem]{ulem}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{mathrsfs}
\usepackage{cleveref}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
%\newdefinition{definition}{Definition}
%\newdefinition{example}{Example}
%\newproof{proof}{Proof}

@@

\author{Onyekachi Emenike, Fred J. Hickernell, Peter Kritzer}
%\ead{kachi.emenike12@gmail.com}
%\author[1]{Fred J. Hickernell}
%\ead{hickernell@iit.edu}
%\author[2]{Peter Kritzer}
%\ead{peter.kritzer@oeaw.ac.at}

\date{\today}
%\affiliation[1]{organization={Department of Applied Mathematics, Illinois Institute of Technology},
%addressline={10 West 32${}^{\text{nd}}$ Street, Room 220},
%postcode={60616},
%city={Chicago, IL},
%country={United States}}
%\affiliation[2]{organization={Johann Radon Institute for Computational and Applied Mathematics (RICAM), Austrian Academy of Sciences},
%addressline={Altenbergerstr. 69},
%postcode={4040},
%city={Linz},
%country={Austria}}
\maketitle

@@

%\maketitle

@@

We give the proof of Proposition \ref{prop:equiv_conditions} here.
\begin{proof}
%[Proposition 1]
We will give the proof of equivalence of the conditions \eqref{eq:exp_pt} and \eqref{eq:KW_exp_pt}. The proof of the equivalence of \eqref{eq:exp_spt} and \eqref{eq:KW_exp_spt} is analogous and simpler (actually, the latter proof can be seen as a special case of the former, by assuming all exponents of $d$ to be zero).
